A Large-scale Study of Polymorphism in R

### 2019-03-30 paper/main.tex
working L1 into sec:results

@@

	\item vectors!

@@

	\item similarly with NULLs, if we encounter a ...

	\item indices?

@@

\AT{!!! The layout of this section will change drastically. !!!}
\AT{Old stuff beyond.}

\AT{Limitations of L0.}
... but there are some clear limitations to this approach.  For
example, \code{typeof} makes no distinction between vectors and scalars, as
indeed there is no such distinction in R.  But that doesn't mean that there
{\it couldn't} be a distinction, as an all-new runtime environment design
for R might benefit from e.g. not needing to vectorize scalar values.

To address these limitations, we augment the logical naive type system with
more information which was alluded to in the discussion of our data
collection methodology.  We will consider unit-length vectors to be scalars,
collect list member type information to build an accurate list type, and
promote NAs to their own type.  In addition, we collect attribute and class
information to distinguish a type for data frames and matrices, as the
\code{typeof} these is simply {\tt list} in R.  This type system is referred
to as the {\it Fine-Grained Type System}, or FGTS.  An example of a signature in
this type system can be found in Figure~\ref{fig:exFGTS}.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
add_make_list <- function(x, y) {
  l <- as.list(x + y)
  attr(l, "example") <- "this is an attribute"
  l
}
add_make_list(2, c(3, 1)) # => list(5, 3)
\end{lstlisting}}
\begin{tabular}{@{}r|l|l|l@{}}\hline
\bf Argument & \bf Type & \bf Class &\bf Attributes \\
x & \tt scalar/double & \tt numeric & \tt \{\} \\
y & \tt vector/double & \tt numeric & \tt \{\} \\
retv & \tt list<double> & \tt list & \tt \{example: scalar/character\}  
\end{tabular}
\caption{Example call and FGTS signature}\label{fig:exFGTS}\end{figure}

Broadly, the first question we will address is: {\it how polymorphic is R?} 
We will answer this in two parts, and first we will look at the number of monomorphic {\it arguments} across all of the packages we analyzed.

%
%
%
%
\subsection{Argument Polymorphism}

In this section, we will explore the polymorphicity of function argument positions and return values.
Exploring argument polymorphism will help to motivate a design for whole function signatures.

To begin, let's look at the {\it amount} of polymorphism by way of {\it number of polymorphic arguments}.
The data can be found in Tables~\ref{tab:argcountsFGTS}~and~\ref{tab:funcountsFGTS}.
\AT{TODO: have these tables be generated automatically}

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:03:30 2019
\begin{table}[ht]
\label{tab:argcountsFGTS}
\centering
\begin{tabular}{lrr}
  \hline
 Argument Polymorphism & Count & Percentage \\ 
  \hline
  Full Monomorphic & 781K & 85.37 \\ 
  Monomorphic in Type & 834K & 91.23 \\ 
  Monomorphic in Class & 827K & 90.42 \\ 
  Monomorphic in Attribute Pattern & 825K & 90.27 \\ 
  Total Seen & 914K & --- \\ 
     \hline
\end{tabular}
\caption{Account of {\it argument} polymorphism in the FGTS.}
\end{table}

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:10:46 2019
\begin{table}[ht]
\label{tab:funcountsFGTS}
\centering
\begin{tabular}{lrr}
  \hline
 Function Polymorphism & Count & Percentage \\ 
  \hline
  Full Monomorphic & 143K & 65.88 \\ 
  Monomorphic in Type & 167K & 77.25 \\ 
  Monomorphic in Class & 163K & 75.24 \\ 
  Monomorphic in Attribute Pattern & 161K & 74.37 \\ 
  Total Seen & 216K & --- \\ 
   \hline
\end{tabular}
\caption{Account of {\it function} polymorphism in the FGTS.}
\end{table}

\AT{We should figure out the pipeline to get the notebook output into here.}

In the tables, ``Full Monomorphic'' denotes monomorphism in type, class, and attribute.
We see that roughly 9\% of all argument positions in R are polymorphic in type.
The numbers for attribute and class polymorphism are included for
completeness, but the important number is that of type polymorphism (as our
method of constructing types sometimes factors in class and attribute).
Specifically for classes, often the class and \code{typeof} a value will
line up (refer to Section~\ref{sec:rlang}).

@@

These numbers alone are insufficient, so we ask: {\it of the polymorphic arguments, what are the most common patterns?}
The answer to that question can be found in Table~\ref{tab:toppolyFDTS}.

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 21:27:37 2019
\begin{table}[ht]
\label{tab:toppolyFDTS}
\centering
\begin{tabular}{lrr}
  \hline
 Polymorphic Type Signature & \# Occurrences & \% of Total Polymorphism \\ 
  \hline
  vector/double, vector/integer & 7742 & 9.66 \\ 
  scalar/double, scalar/integer & 7201 & 8.99 \\ 
  matrix, vector/double & 5384 & 6.72 \\ 
  NULL, scalar/character & 2325 & 2.90 \\ 
  list$<$any$>$, list$<$double$>$ & 2231 & 2.78 \\ 
  data.frame, matrix & 2116 & 2.64 \\ 
  NULL, scalar/double & 2082 & 2.60 \\ 
  NULL, vector/double & 2069 & 2.58 \\ 
  list$<$any$>$, list$<$list$>$ & 1432 & 1.79 \\ 
  NULL, vector/character & 1380 & 1.72 \\
     \hline
\end{tabular}
\caption{Top polymorphic argument signatures in the FDTS.}
\end{table}

\AT{This should probably be discussed earlier.}
Not pictured here, the most common polymorphic pattern in the FGTS is the {\tt scalar/X}, {\tt vector/X} pattern, but this is dealt with separately:
If a function argument is observed to be both a unit-length and non-unit-length vector, then it's reasonable to assume that the argument is intended to be a vector, so we collapse the distinction in such cases.

In Table~\ref{tab:toppolyFDTS}, we see the prominence of {\tt double} and {\tt integer} polymorphism.
In many dynamic languages, primitive double and integer types can easily stand-in for each other as coercion between them is easy.
In R, the distinction is mainly just relevant to the implementation, as integers and doubles are stored and dealt with differently in the runtime but programmers only see the distinction when printing.
We address this by creating a new type, {\it real}, encompassing both.
The expected coercion is evident throughout R.
Refer to the snippet in Figure~\ref{fig:realex} for details.

@@

Another clear pattern in Table~\ref{tab:toppolyFDTS} is that of NULL, X.
In many programming languages, NULLs inhabit every type, and this is not the case in our FGTS.
The FGTS reflects many of R's implementation details, and this distinction is no exception:
NULLs are represented as \AT{fill in} in the R runtime.

@@

Not pictured in Table~\ref{tab:toppolyFDTS} are NAs, which are ascribed the type {\tt raw\_NA} in FGTS; NAs in fact inhabit every type, though the default NA type is {\tt logical}.
Just as we fold NULL into all types, we will do the same with NAs.

@@

%\item {\it The Fine-Grained Type System (FGTS)}: this type systems includes a number of relevant additions.

@@

%Of course, one could conceive of a more fine-grained type system than FGTS, but it's unclear that we would gain much in doing so.

@@

%FGTS distinguishes such NAs with the {\tt raw\_NA} type, and if instead we consider NA to inhabit all types we may \AT{win}.

>>>>

%
%
%
%
\subsection{L0}

@@

	\item to get an idea of how often vectors and lists co-occur in signatures, we need to expand our analysis of vectorized types.
	Notably, when we encounter e.g. a double, we should collect length information to see if the double is a vector of doubles or just a scalar.
	This will feature in our next type system, L1, which we will get to momentarily.

@@

	\item similarly with NULLs, we're interested in rolling NAs in to all types.
	As we mentioned, this is actually what the R runtime does, but a default type needed to be chosen for NAs appearing without context.
	Subsuming NAs should help to reveal other interesting patterns.

	\item dealing with this polymorphism will be more tricky.
	We might need to pull class information in (to pick up on dates), and possibly make a signature judgment depending on which argument position has the signature (e.g. a \{character, X\} signature on a return).
	We will come back to this later.

@@

There are some clear limitations to the L0 type system.
As we mentioned, \code{typeof} makes no distinction between vectors and scalars, as indeed there is no such distinction in R. 
But as we saw, we might benefit from such a distinction, and indeed an all-new runtime environment design for R might benefit from e.g. not needing to vectorize scalar values.

%
%
%
%
\subsection{L1}

To address these limitations, we augment the L0 with more fine-grained information.
We will consider unit-length vectors to be scalars, collect list member type information to build an accurate list type, and promote NAs to their own type. 
In addition, we collect attribute and class information to distinguish a type for data frames and matrices, as the \code{typeof} these is simply {\tt list} in R. 
We refer to this as the L1 type system.  An example of a signature in this type system can be found in Figure~\ref{fig:exL1}.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
add_make_list <- function(x, y) {
  l <- as.list(x + y)
  attr(l, "example") <- "this is an attribute"
  l
}
add_make_list(2, c(3, 1)) # => list(5, 3)
\end{lstlisting}}
\begin{tabular}{@{}r|l|l|l@{}}\hline
\bf Argument & \bf Type & \bf Class &\bf Attributes \\
x & \tt scalar/double & \tt numeric & \tt \{\} \\
y & \tt vector/double & \tt numeric & \tt \{\} \\
retv & \tt list<double> & \tt list & \tt \{example: scalar/character\}  
\end{tabular}
\caption{Example call and L1 signature.}\label{fig:exL1}\end{figure}

Let us begin as we did in L0 by looking at the amount of polymorphism in L1.
The data can be found in Tables~\ref{tab:argcountsL1}~and~\ref{tab:funcountsL1}.
\AT{TODO: have these tables be generated automatically}

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:03:30 2019
\begin{table}[ht]
\label{tab:argcountsL1}
\centering
\begin{tabular}{lrr}
  \hline
 Argument Polymorphism & Count & Percentage \\ 
  \hline
  Full Monomorphic & 781K & 85.37 \\ 
  Monomorphic in Type & 834K & 91.23 \\ 
  Monomorphic in Class & 827K & 90.42 \\ 
  Monomorphic in Attribute Pattern & 825K & 90.27 \\ 
  Total Seen & 914K & --- \\ 
     \hline
\end{tabular}
\caption{Account of {\it argument} polymorphism in L1.}
\end{table}

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:10:46 2019
\begin{table}[ht]
\label{tab:funcountsL1}
\centering
\begin{tabular}{lrr}
  \hline
 Function Polymorphism & Count & Percentage \\ 
  \hline
  Full Monomorphic & 143K & 65.88 \\ 
  Monomorphic in Type & 167K & 77.25 \\ 
  Monomorphic in Class & 163K & 75.24 \\ 
  Monomorphic in Attribute Pattern & 161K & 74.37 \\ 
  Total Seen & 216K & --- \\ 
   \hline
\end{tabular}
\caption{Account of {\it function} polymorphism in L1.}
\end{table}

\AT{We should figure out the pipeline to get the notebook output into here automatically.}

We see that roughly 9\% of all argument positions in R are polymorphic in type.
The numbers for attribute and class polymorphism are included for
completeness, but the important number is that of type polymorphism (as our
method of constructing types sometimes factors in class and attribute).
Specifically for classes, often the class and \code{typeof} a value will
line up (refer to Section~\ref{sec:rlang}).

@@

These numbers alone are insufficient, so we ask: {\it of the polymorphic arguments, what are the most common patterns?}
The answer to that question can be found in Table~\ref{tab:toppolyL1}.

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 21:27:37 2019
\begin{table}[ht]
\label{tab:toppolyL1}
\centering
\begin{tabular}{lrr}
  \hline
 Polymorphic Type Signature & \# Occurrences & \% of Total Polymorphism \\ 
  \hline
  vector/double, vector/integer & 7742 & 9.66 \\ 
  scalar/double, scalar/integer & 7201 & 8.99 \\ 
  matrix, vector/double & 5384 & 6.72 \\ 
  NULL, scalar/character & 2325 & 2.90 \\ 
  list$<$any$>$, list$<$double$>$ & 2231 & 2.78 \\ 
  data.frame, matrix & 2116 & 2.64 \\ 
  NULL, scalar/double & 2082 & 2.60 \\ 
  NULL, vector/double & 2069 & 2.58 \\ 
  list$<$any$>$, list$<$list$>$ & 1432 & 1.79 \\ 
  NULL, vector/character & 1380 & 1.72 \\
     \hline
\end{tabular}
\caption{Top polymorphic argument signatures in the FDTS.}
\end{table}

\AT{This should probably be discussed earlier.}
Not pictured here, the most common polymorphic pattern in L1 is the {\tt scalar/X}, {\tt vector/X} pattern, but this is dealt with separately:
If a function argument is observed to be both a unit-length and non-unit-length vector, then it's reasonable to assume that the argument is intended to be a vector, so we collapse the distinction in such cases.

In Table~\ref{tab:toppolyL1}, we see the prominence of {\tt double} and {\tt integer} polymorphism.
In many dynamic languages, primitive double and integer types can easily stand-in for each other as coercion between them is easy.
In R, the distinction is mainly just relevant to the implementation, as integers and doubles are stored and dealt with differently in the runtime but programmers only see the distinction when printing.
We address this by creating a new type, {\it real}, encompassing both.
The expected coercion is evident throughout R.
Refer to the snippet in Figure~\ref{fig:realex} for details.

@@

Another clear pattern in Table~\ref{tab:toppolyFDTS} is that of NULL, X.
In many programming languages, NULLs inhabit every type, and this is not the case in our L1.
L1 reflects many of R's implementation details, and this distinction is no exception:
NULLs are represented as \AT{fill in} in the R runtime.

@@

Not pictured in Table~\ref{tab:toppolyFDTS} are NAs, which are ascribed the type {\tt raw\_NA} in L1; NAs in fact inhabit every type, though the default NA type is {\tt logical}.
Just as we fold NULL into all types, we will do the same with NAs.

@@

%\item {\it The Fine-Grained Type System (L1)}: this type systems includes a number of relevant additions.

@@

%Of course, one could conceive of a more fine-grained type system than L1, but it's unclear that we would gain much in doing so.

@@

%L1 distinguishes such NAs with the {\tt raw\_NA} type, and if instead we consider NA to inhabit all types we may \AT{win}.
